---
title: "Regex Problems"
author: "Ben Schmidt"
output: pdf_document
---

## Regex Exercises

```{r, include = FALSE}
library(tidyverse)
library(HumanitiesDataAnalysis)
knitr::opts_chunk$set(eval=FALSE)
```

**Try** to think about these. Give it a shot. Some of them are hard--if you're simply banging your head against the wall, take a break, or simply don't finish.

### Searches

1. There is one word that ends with two 'u's. What is it?

```{r}
dictionary_search("YOUR SEARCH HERE")
```

2. The word "picalilli" contains five consecutive "l" or "i" letters. What word contains 6 consecutive "i" or "ls"?

```{r}
dictionary_search("SEARCH")
```

3. What is the longest substring of your name for which a word contains all the matches, in order? For example, my name is "Ben Schmidt" and I can match the first five letters wtih the capitalized letters in the word `BirkENStoCk.` What is yours?

```{r}


```


4. What dictionary words contain the same letter, three times in a row? [hard]

```{r}
dictionary_search("SEARCH")
```

5. Besides the word found in question 2, are there any other words in the
dictionary that contain two identical letters 6 times in a row? 7
times in a row? [harder]

```{r}


```

### Replacements

6. One way that online sites occasional deal with abusive posts is to 'disemvowel' them; that is, to remove all of the vowels from them. This requires a replacement regular expression.

```{r}
str_replace("", "PATTERN", "REPLACEMENT")
```

7. Sometimes documents have excessive spaces in them. (For instance,
if you copy and paste from the Internet). Write a regex
that reduces any string of spaces down to just one. **For example,** this text:

```
	Good day, everyone.
	1     4      6
	
	3   4       10
	Good night, ladies
```

Would be reduced to:

```
Good day, everyone. 1 4 6 3 4 10 Good night, ladies 
```
	
```{r space_replacement}

str_replace("This   is    too
            many spaces
            
            Isn't it?", "PATTERN", "REPLACEMENT")

```

8. Write a regex that changes the spelling of all words in a document
so that they conform to the rule "I before e, except after c." [hard]

```{r}

```

*Concordances*

An online version of the bible is at
benschmidt.org/bible. It allows you to filter and replace at
once on the bible. This may take some time to run, so it initially
will only show values for the book of Matthew.

9. Pick a biblical figure--Jesus, for instance. Assume that the Bible is in the past tense, and that all verbs end with 'ed'. What sort of 
stuff does your figure do the most? What are his most common verbs? (You don't have to use the replacements, but there's a way to copy and paste the top line and use a single `\1` grouping
that will make the output much easier to read.)

10. Create a regex that reduces the bible to a concordance for the word "love" that shows 3 words before and 3 words after. (very hard)

11. Edit that regex so that it includes the book/line/verse number as the beginning. (very hard)
