---
title: "Working in R"
author: "Ben Schmidt"
date: "2019-01-17"
output: pdf_document
...

# Working in R

## Some history

Every computer language is an accretion of cultural history; knowing a little bit about R will help you to understand what's happening in it.

R, the programming language, dates back to the 1970s. During the heyday of Bell Labs in the 1970s,
researchers built a variety of tools for working with different computer systems, including the language `C` that has influenœced most low-level program design since, and the operating system `Unix` that provides the foundation for many modern computing systems from Apple laptops to Amazon servers to Android phones.

John Chambers developed a language called "S" at Bell Labs with several goals that continue to influence the language's design. One was to provide a way to use, in a more human notation, the blazingly fast linear algebra routines that undergird all sorts of work in math, statistics, and visualization. Another was to facilitate more sophisticated, exploratory data visualization. 

In the 1990s, two statisticians in New Zealand, Ross Ihaka and Robert Gentleman, created an open source version of S called "R" that could be freely distributed without worrying about AT&T's old patents. By virtue of being free, that language has slowly displaced Stata and SPSS, the other major statistical computing environments of the 1980s and 1990s.

## Packages

R is a modular piece of software. The base language allows you to apply most standard statistical methods from the 1990s, but in general the power of the language comes from extensions that others have worked with. This means that even after installing R, you'll frequently have to install additional "packages." The ability to do so is contained inside R itself; you can think of packages as living **inside** your local copy of R.

There is also one program worth installing that lives outside, called "RStudio."

### The **`Tidyverse`**

One set of packages bears particular emphasis. Hadley Wickham, who was for a time a statistics professor at Rice University, and more recently has become the chief data scientist at RStudio. The `tidyverse` is a set of packages that he oversees that provides a different syntax for R.

### Installing from CRAN

R packages can come from two sources. The first, easier one is internal to the R ecosystem and called "CRAN" (The 'Comprehensive R Archive Network'). 
CRAN sets the highest bar on what packages are available.

The function **`install.packages`** or the "packages" pane in RStudio.

```{r eval=FALSE}

install.packages("tidyverse")

install.packages("remotes")


```

### Installing from github

But it can be a great deal of work to make a package fit into CRAN--the maintainers are famously fastidious about certain standards that can be included. (The digital humanist Matt Lincoln has a blog post about how an obscure feature of certain operating systems nearly broke not just his clipoard package, but all sorts of other packages that depend on it. @lincoln_little_2019) Frequently you'll want to install packages from outside sources; the most common is Github, a website owned by Microsoft that distributes code using the open source 'git' standard.


### The course package

This course itself uses an R package. This gives you a place from which to download 

```{r eval = FALSE}
remotes::install_github("HumanitiesDataAnalysis/HumanitiesDataAnalysis")
```
