---
title: "Grouping, filtering, and counting"
author: "Ben Schmidt"
date: "2019-01-29"
output: pdf_document
---

Week 3 Problem set.

## Crews dataset.

1. Using summarize, find the minimum age in the set. (Remember, the function to find a minimum is `min`.)

```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(eval=FALSE, warning=FALSE, message=FALSE)
```

``` {r Exercise 1a}
library(tidyverse)
data(crews)

```

2. Using `filter`: what is the name of that youngest person? When did he or she sail?

``` {r Exercise 2a}
?set_tidy_names
```


3. How many sailors left on Barks between 1850 and 1880? Chain together `filter` and `summarize` with the special `n()` function. Note that this has a number of different conditions in the filter statement. You could build several filters in a row: but you can also include multiple filters by separating them with commas. For instance, `filter(school=="Northeastern",year==2015)` might be a valid filter on some dataset (though not this one.)


``` {r Exercise 3a}

```

4. Question 3 told you how many sailors left on barks in those years. How many distinct voyages left? The variable `Voyage.number` identifies distinct voyages in this set. (This may require reading some documentation: reach out to me or a classmate if you can't figure it out. There are at least two ways: one involves using the `dplyr` function `distinct` before summarizing, and the second involves using the functions `length` and `unique` in your call to `summarize`.)

``` {r Exercise 4a}

```

## Grouping

``` {r Group function}
crews %>% group_by(Residence)
```

Now when we run a summary operation, it will give the counts. For example, try counting the number of people by residence.

``` {r Grouping example}
crews %>% group_by(Residence) %>%
```




### String functions

To think about working with *text*, I'm going to introduce a new dataset that I think might be interesting: all the books in the Library of Congress catalog that have summaries. There's a larger set I can supply, but these are further reduced down to "just" about 150,000 books that have dates of birth for their authors.

The `tidyverse` packages includes some functions for exploring strings we can use inside these functions.

`str_detect` *finds* a regular expression; it returns TRUE or FALSE, so works well with `filter`.
`str_replace` *replaces* a regular expression inside a string: it returns a new string, so it works well with mutate.
`str_length` returns a number which is the *length* of the summary.


```{r}

books %>% 
  filter(subjects %>% str_detect("France|French")) %>% 
  select(year, summary) %>% 
  sample_n(10)

```


```{r}
country_regex = ".*(France|Germany|United Kingdom).*"
books %>% 
  filter(subjects %>% str_detect(country_regex)) %>%
  mutate(
  country = subjects %>% str_replace(country_regex, "\\1")
) %>% filter(!is.na(country)) %>%
  group_by(country) %>% 
  summarize(count=n())
```

