---
title: Cleaning Data
author: Benjamin Schmidt
date: 2019-01-19
output: pdf_document
---


## Exercises: tidying data

1. Is the data you read in under shiptypes `tidy`?

2. Using the `gather` and `spread` functions from the `tidyr` package, create a new data.frame that has rows corresponding to ship types and columns corresponding to years. (That is, the exact inverse of the data.frame you saw before.)

``` {r}



```

3. There's another file in the `data` directory called CESTACityData.csv. It was graciously provided by the [Center for Spatial and Textual Analysis at Stanford](https://cesta.stanford.edu/). Read it in. We'll be working with this data set more next week.

``` {r}
CESTA
```


4. This data is clearly not tidy. (Why not?) Use `gather` to turn it into tidy data, and save it as a data.frame called "tidied". (Hint: the final frame will have a column called "year:" You will have to use gsub to remove Xs from the year name.)

``` {r}

```

5. Use the `?` operator to read about the function `write_csv`. Save the result to disk: we'll be exploring it more next week.

``` {r}

```


### Tidy data for network visualizations.

If you are interested in network visualizations and have data like So and Long's, come to me: we'll talk about how to use tidyr to reshape some data for network analysis to be exported into a program like Gephi.
